---
title: "Sjogren Syndrome - ITC on TRAE"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-tools: true
    code-summary: "Show/Hide Code"
knitr:
  opts_chunk:
    fig.path: "figures/TRAE/"
---

## Required Packages

```{r}
#| message: false
#| warning: false
library(openxlsx)
library(gt)
library(gemtc)
library(rjags)
library(tidyverse)
```

## Loading Data

```{r}
df0 <- read.xlsx("Safety_Outcomes.xlsx", sheet = "TRAE")
```

```{r}
#| include: false
df0 |> gt()
```

```{r}
df1 <- data.frame(study = df0$STUDY,
                  treatment = df0$TREATMENT,
                  responders = df0$N.EVENT,
                  sampleSize = df0$N)
# column names MUST be these

df1 |> gt()
```

## Creating the Network

```{r}
network <- mtc.network(data.ab = df1)
```

```{r}
summary(network)
```

## Network Diagram

```{r}
plot(network)
```

## Model Fitting

### Building the NMA Model

`mtc.model()` generates random numbers using the R functions. Therefore, `set.seed()` is called before calling `mtc.model()`.

```{r}
set.seed(14)
```

```{r}
model <- mtc.model(network,
                   likelihood = "binom",
                   link = "log",
                   linearModel = "fixed",
                   om.scale = NULL,
                   n.chain = 4)
```

`om.scale` stands for Outcome Magnitude scale and it is used to set the width of essentially every "vague" prior in the model. Prior to heterogeneity parameter is set to `dunif(0, om.scale)` and priors of the treatment effects are set to `N(0, sd = 15 * om.scale)`.

### Running MCMC

Before calling `mtc.run()`, the random number generator type and seed are set for each of the chains. Refer to [gemtc::mtc.model documentation](https://cran.r-project.org/web/packages/gemtc/refman/gemtc.html#mtc.model) for more.

```{r}
seeds <- sample.int(4, n = .Machine$integer.max)

model$inits <- mapply(c, model$inits, list(
  list(.RNG.name="base::Wichmann-Hill", .RNG.seed=seeds[1]),
  list(.RNG.name="base::Marsaglia-Multicarry", .RNG.seed=seeds[2]),
  list(.RNG.name="base::Super-Duper", .RNG.seed=seeds[3]),
  list(.RNG.name="base::Mersenne-Twister", .RNG.seed=seeds[4])), 
  SIMPLIFY=FALSE)
```

Wichmann-Hill, Marsaglia-Multicarry etc. are the names of different pseudo-random number generator (PRNG) algorithms built into JAGS. JAGS needs each MCMC chain to use an independent stream of random numbers, so it ships with several distinct generator algorithms - a different one is assigned to each chain so their random draws don't overlap or correlate.

```{r}
results <- mtc.run(model, 
                   n.adapt = 5000, 
                   n.iter  = 20000)
```

### Assessing MCMC Convergence

In case of convergence, Potential Scale Reduction Factor (PSRF) should gradually shrink down to $0$ with increasing numbers of iterations, and should at least be below $1.05$ in the end.

```{r}
gelman.plot(results)
```

From the plots, we can be assured of convergence.

## Posterior Summaries

```{r}
summary(results)
```

## Relative Effects

```{r}
Relative.Effects <- relative.effect.table(results)
Relative.Effects
```

The above results are in log-RR scale. Let us have them in RR scale.

```{r}
exp(Relative.Effects)
```

## Final Result

Extract the necessary information.

```{r}
RR.array <- exp(Relative.Effects)

get_comparison <- function(arr, new_drug, reference) {
  data.frame(
    Comparison = paste(new_drug, "vs.", reference),
    RR = round(arr[reference, new_drug, "50%"], 2),
    Lower = round(arr[reference, new_drug, "2.5%"], 2),
    Upper = round(arr[reference, new_drug, "97.5%"], 2)
  )
}

RR_table <- rbind(
  get_comparison(RR.array, "VAY736QM", "Placebo"),
  get_comparison(RR.array, "Telitacicept_160mg", "Placebo"),
  get_comparison(RR.array, "VAY736QM", "Telitacicept_160mg")
)

rownames(RR_table) <- NULL
```

```{r}
data.frame(RR_table) |> gt()
```

## Forest Plot

```{r}
#| include: false
df1 <- data.frame(comparison = gsub("_", " ", RR_table$Comparison),
                  RR = RR_table$RR,
                  lower = RR_table$Lower,
                  upper = RR_table$Upper)
```

```{r}
#| include: false
df2 <- df1 %>%
  mutate(
    Significant = dplyr::if_else(lower < 1 & upper < 1, "Yes", "No"),
    Label = sprintf("%.2f (%.2f, %.2f)", RR, lower, upper)
  ) %>%
  mutate(comparison = factor(comparison, levels = unique(comparison)))
  # Preserve the order of treatments as they appear in the data
```

```{r}
#| include: false
forest_plot <- df2 %>% ggplot(
  aes(x = RR,
      y = comparison,
      xmin = lower,
      xmax = upper,
      colour = Significant)) +
  geom_errorbar(orientation = "y", width = 0.25, linewidth = 0.9, colour = "#2166AC") +
  geom_point(size = 4, shape = 18, colour = "#2166AC") +
  geom_vline(xintercept = 1, linetype = "dashed", 
             colour = "grey40", linewidth = 0.7) +
  geom_text(aes(x = upper, label = Label),
            hjust = -0.08, size = 3.2, colour = "grey20") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.35))) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey40", size = 10),
        plot.caption = element_text(colour = "grey50", size = 9),
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(size = 11))
```

```{r out.width="100%"}
#| echo: false
#| message: false
#| warning: false
forest_plot
```

```{r}
#| message: false
#| warning: false
#| include: false
ggsave("Forest_Plots/FP_TRAE.png",
       forest_plot,
       width = 10.5,
       height = 5.5,
       dpi = 2000,
       bg = "white", 
       create.dir = TRUE)
```
