---
title: "Bayesian Random-effects Meta Regression"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
format: html
editor: visual
knitr:
  opts_chunk:
    fig.path: "figures/"
---

## Required Libraries

```{r message=FALSE, warning=FALSE}
library(openxlsx)

library(gt)
library(tidyverse)

library(cmdstanr)

library(bayesplot)
```

## Loading Data

```{r}
#| include: false
data.path <- 'https://github.com/ananda-ebm/datasets/raw/refs/heads/master/NSCLC.xlsx'
```

```{r}
df1 <- read.xlsx(data.path, sheet = 1)
```

## Initial Sanity Checks

```{r}
str(df1)
```

```{r}
dim(df1)
```

```{r, results='asis'}
#| echo: false
#| tbl-cap: "Snapshot of NSCLC Data"

head(df1) |>
  gt() |>
  tab_options(
    data_row.padding = px(8),
    column_labels.padding = px(10)
  )
```

## Derived Quantities

```{r}
df1$weight <- 1 / (df1$OS.se)^2
```

## Generating Treatment Effect on Surrogate Endpoint in New Trials

```{r}
range(df1$HR.PFS)
```

```{r}
new.data <- data.frame(PFS.ln.HR = log(seq(0.2,3, length=1000)))
```

## Model

A random-effects meta-analysis allows for the differences in the treatment effect between studies. In a Bayesian random-effects meta-regression framework, the model can be written as

$$Y_{2i}
\sim
N\left(\mu_{2i},\sigma_{2i}^2\right),
\qquad
i=1,\ldots,N,$$

where the mean (study-specific treatment effect) is modeled as

$$\mu_{2i}
=
\lambda_{0i}
+
\lambda_1 Y_{1i}.$$

The study-specific intercepts are assumed to follow a common Normal distribution:

$$\lambda_{0i}
\sim
N\left(\beta,\psi^2\right).$$

Here:

- $Y_{1i}$ = summary measure of the treatment effect on the candidate surrogate outcome in study $i$;
- $Y_{2i}$ = summary measure of the treatment effect on the target outcome in study $i$;
- $\mu_{2i}$ = underlying true treatment effect on the target outcome in study $i$;
- $\sigma_{2i}$ = standard deviation of the estimated treatment effect $Y_{2i}$;
- $\lambda_{0i}$ = study-specific true effect on the target outcome when the treatment effect on the surrogate is zero;
- $\lambda_1$ = regression coefficient describing the relationship between treatment effects on the surrogate and target outcomes;
- $\beta$ = mean of the overarching distribution of the study-specific intercepts; and
- $\psi^2$ = between-study variance representing heterogeneity in the study-specific intercepts.

In the Bayesian framework, all parameters are given prior distributions :

$$\beta \sim N(0,10^2),$$

$$\lambda_1 \sim N(0,10^2),$$

$$\psi \sim \operatorname{Half}\text{-}t_3(0,2.5), \qquad \psi > 0,$$

and

$$\lambda_{0i}^{\mathrm{raw}} \sim N(0,1),
\qquad i=1,\ldots,N.$$

## Model Fitting

### Preparing data for STAN

```{r}
STAN_data.BREMR <- list(
  N = nrow(df1),
  Y2 = df1$OS.ln.HR,
  Y1 = df1$PFS.ln.HR,
  sigma2 = df1$OS.se
)
```

### STAN Model Compilation

```{r}
#| include: false
BREMR.model.path <- "C:\\Users\\AnandaBiswas\\OneDrive - EBM Health Consultants LLP\\Documents\\R-Programs\\Self-Study\\0021_Comparison_of_Surrogacy_Models_on_NSCLC_data\\0021__03_Bayesian_Random-effects_Meta_Regression\\0021__03_BREMR_model.stan"
```

```{r}
model.BREMR <- cmdstan_model(stan_file = BREMR.model.path)
```

### Running HMC

```{r}
n.chains <- 4
n.samples.per.chain <- 10000
```

```{r}
fit.BREMR <- model.BREMR$sample(
  data = STAN_data.BREMR,
  chains = n.chains,
  parallel_chains = n.chains,
  iter_warmup = 1000,
  iter_sampling = n.samples.per.chain,
  seed = 14,
  refresh = 0 # suppress the iteration progress messages
)
```

```{r}
n.posterior.samples <- n.chains * n.samples.per.chain
```

### STAN Diagnostics

```{r}
fit.BREMR$cmdstan_diagnose()
```

### Posterior Predictive Check - PPC

It is a way of checking whether my Bayesian model can generate data that look reasonably similar to the data I actually observed.

Suppose we observe $n$ data-points. For each posterior draw of the parameters, we draw a random sample of $n$ data-points.

Here in NSCLC data-set, there are $`r nrow(df1)`$ studies. So we shall have $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ replicated data-sets, each having $`r nrow(df1)`$ studies.

```{r}
Y2.replicated.draws <- fit.BREMR$draws(variables = "Y2_rep", format = "matrix")

dim(Y2.replicated.draws)
```

Plot the density of the observed data and densities of a few randomly selected replicated data-sets to assess the Bayesian model.

```{r fig.height=6, fig.width=8, out.width="100%"}
#| message: false
#| warning: false
# a few randomly selected replications
temp <- sample(1:n.posterior.samples, 300)

ppc_dens_overlay(y = df1$OS.ln.HR,
                 yrep = Y2.replicated.draws[temp, ])
```

Plot the observed `y~x` along with the confidence intervals calculated from the replications of each `y`.

For instance, here I shall have $`r nrow(df1)`$ points each corresponding to a study in the NSCLC data, and $`r nrow(df1)`$ confidence intervals accompanying them. Confidence interval for $Y_{2j}$ was calculated from the $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ replications of $Y_{2j}$ $\forall j = 1(1)`r nrow(df1)`$.

```{r fig.height=6, fig.width=8, out.width="100%"}
#| message: false
#| warning: false
ppc_intervals(y = df1$OS.ln.HR,
              yrep = Y2.replicated.draws,
              x = df1$PFS.ln.HR)
```

## Posterior Summaries of Parameters of Interest

```{r}
fit.BREMR$summary(variables = c("beta", "lambda1", "psi")) |>
  gt()
```

Posterior means/medians can be used as Bayesian estimates of the parameters.

## Prediction of TE on Clinical Outcome in New Trials

Now I take posterior sample to calculate the predictions on new future trials, credible intervals and prediction intervals. These are the main components of **trial-level association plot**.

```{r}
posterior.draws <- fit.BREMR$draws(
  variables = c("beta", "lambda1", "psi"),
  format = "matrix"
)
```

```{r}
#| tbl-cap: "Preview of Posterior Sample"

data.frame(head(posterior.draws)) |>
  gt()
```

```{r}
dim(posterior.draws)
```

Each row of `posterior.draws` is a posterior sample from the joint posterior distribution of the parameters.

Now I create a design matrix `X` representing new studies on which we will make future predictions.

```{r}
X <- matrix(c(rep(1, 1000), new.data$PFS.ln.HR), 
            nrow = 1000, ncol = 2, byrow = FALSE)
```

For each new trial in `X`, corresponding to $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ posterior samples of the parameters, there will be $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ predictions of `log.HR.OS`. Below I do a matrix operation to obtain all the $1000 \times `r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ predictions.

```{r}
BREMR.posterior <- X %*% t(posterior.draws[,1:2])
```

```{r}
dim(BREMR.posterior)
```

Each row of `BREMR.posterior` represents $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ posterior draws of `log.HR.OS` *i.e.* the response variable, corresponding to a new `log.HR.PFS` of a new trial.

### Point Estimates

For a new trial, to get a prediction of `log.HR.OS`, we must calculate the mean of the posterior samples.

```{r}
BREMR.fit <- apply(BREMR.posterior, 1, mean)
```

`BREMR.fit` has $1000$ predictions corresponding to $1000$ new future trials.

### Credible Intervals

Credible intervals are obtained from the quantiles of the posterior distribution. Here I calculate the $95\%$ credible intervals for the predictions on new trials.

```{r}
BREMR.ci <- t(apply(BREMR.posterior, 1, quantile, probs = c(0.025, 0.975)))
```

### Prediction Intervals

Prediction intervals are obtained from the quantiles of the posterior predictive distribution.

The posterior distribution of concern here is

$$p\left(\beta, \lambda_1, \psi \mid Y_{11}, Y_{21}, Y_{12}, Y_{22}, \ldots, Y_{1N}, Y_{2N}\right).$$ Already I have $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ samples from this posterior distribution.

We want the predictive distribution of $\widehat{\mu_{2 \cdot \text{new}}}:$ the true treatment effect on clinical outcome in a future trial.

In any future trial, let $Y_{1 \cdot \text{new}}$ be the observed treatment effect on surrogate endpoint. Then we must have

$$\widehat{\mu_{2 \cdot \text{new}}} \sim \mathcal{N}\left( \beta + \lambda_1 Y_{1 \cdot \text{new}}, \psi^2\right).$$

In Bayesian theory, the predictive distribution is obtained by averaging the likelihood over the entire parameter space where the parameters vary in accordance to the posterior distribution.

Mathematically, we can write the posterior predictive distribution as

$$\displaystyle{\int \limits_{(\beta, \lambda_1, \psi)}} \mathcal{N}\left( \beta + \lambda_1 Y_{1 \cdot \text{new}}, \psi^2\right) \cdot p\left(\beta, \lambda_1, \psi \mid \cdots \right) d\beta \; d\lambda_1 \; d\psi.$$

As analytical solution cannot be traced, we resort to numerical solution as follows :

- Using posterior sample $(\beta_{(m)}, \lambda_{1(m)}, \psi_{(m)})$ calculate $\beta_{(m)} + \lambda_{1(m)} Y_{1 \cdot \text{new}}$ $\forall m = 1(1)`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$.
- Take a random sample from $\mathcal{N}\left(\beta_{(m)} + \lambda_{1(m)}, \psi_{(m)}^2 \right)$ $\forall m = 1(1)`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$.
- Calculate $q_{0.025}$, $q_{0.975}$ for the random samples just obtained.
- Repeat the above $3$ for all the new trials to get predictive interval around their predictions.

Now I proceed to calculate the $95\%$ posterior prediction intervals.

```{r}
u <- rnorm(n.posterior.samples, 0, sd = posterior.draws[,3])

BREMR.predictions <- BREMR.posterior + 
                     matrix(rep(u, 1000), 
                            nrow = 1000, 
                            ncol = n.posterior.samples,
                            byrow = TRUE)
```

```{r}
BREMR.pi <- t(apply(BREMR.predictions, 1, quantile, probs = c(0.025, 0.975)))
```

## Association Plot

I combine all the required components for the trial-level association plot.

```{r}
BREMR.pred.df <- data.frame(HR.OS = exp(BREMR.fit),
                            HR.PFS = exp(new.data$PFS.ln.HR),
                            lower.ci = exp(BREMR.ci[,1]),
                            upper.ci = exp(BREMR.ci[,2]),
                            lower.pi = exp(BREMR.pi[,1]),
                            upper.pi = exp(BREMR.pi[,2]))
```

```{r}
BREMR.association.plot <- ggplot(df1, aes(x = HR.PFS, y = HR.OS)) +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.ci, ymax = upper.ci), data = BREMR.pred.df, alpha = 0.6, fill = "steelblue") +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.pi, ymax = upper.pi), data = BREMR.pred.df, alpha = 0.4, fill = "lightblue") +
  geom_line(aes(x = HR.PFS, y = HR.OS), data = BREMR.pred.df, 
            col = "steelblue", linewidth = 1, stat = 'identity') +
  geom_point(aes(size = weight), col = "gray30", fill = "orange", shape = 21, data = df1) +
  scale_x_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  scale_y_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  xlab("Progression-free survival hazard ratio (log scale)")+ 
  ylab(paste("Overall survival hazard ratio" , "\n", "\U00B1", "95% confidence and prediction intervals (log scale)")) +
  ggtitle("Bayesian Random-effects Meta-regression") +
  theme_bw() +
  scale_colour_hue("trial") +
  theme(legend.position = "none")
```

```{r, fig.width=8, fig.height=6, out.width="100%"}
BREMR.association.plot
```

## Surrogate Threshold Effect

```{r}
candidate.HR.PFS <- BREMR.pred.df %>%
  filter(upper.pi < 1) %>%
  pull(HR.PFS)

BREMR.STE <- round(max(candidate.HR.PFS), digits = 3)
BREMR.STE
```
