---
title: "Daniels and Hughes Bayesian Bivariate Meta Analysis"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-tools: true
    code-summary: "Show/Hide Code"
knitr:
  opts_chunk:
    fig.path: "figures/"
---

```{r}
#| message: true
#| warning: true
#| include: false
options(mc.cores = parallel::detectCores())
```

## Required Libraries

```{r message=FALSE, warning=FALSE}
library(openxlsx)

library(gt)
library(tidyverse)

library(cmdstanr)
```

## Loading Data

```{r}
#| include: false
data.path <- 'https://github.com/ananda-ebm/datasets/raw/refs/heads/master/NSCLC.xlsx'
```

```{r}
df1 <- read.xlsx(data.path, sheet = 1)
```

## Initial Sanity Checks

```{r}
str(df1)
```

```{r}
dim(df1)
```

```{r, results='asis'}
#| echo: false
#| tbl-cap: "Snapshot of NSCLC Data"

head(df1) |>
  gt() |>
  tab_options(
    data_row.padding = px(8),
    column_labels.padding = px(10)
  )
```

## Derived Quantities

```{r}
df1$weight <- 1 / (df1$OS.se)^2
```

## Generating Treatment Effect on Surrogate Endpoint in New Trials

```{r}
range(df1$HR.PFS)
```

```{r}
new.data <- data.frame(PFS.ln.HR = log(seq(0.2,3, length = 1000)))
```

## Model

Consider a meta-analysis of $N$\$N\$ clinical trials each of which, for simplicity, involves the comparison of two treatments.

In the $i-$th trial,

- $\theta_i$ denotes the true treatment difference on the clinical outcome (true endpoint),
- $\gamma_i$ denotes the true treatment difference on the response variable (surrogate endpoint).

We assume that we have estimates $\widehat{\theta_i}$ and $\widehat{\gamma_i}$, respectively, available from each trial and that the size of each study is sufficiently large, that $(\widehat{\theta_i}, \widehat{\gamma_i})$ is normally distributed about $(\theta_i, \gamma_i)$.

Within the $i-$th clinical trial, we assume that the following model applies:

$$\begin{pmatrix}
\hat{\theta}_i\\
\hat{\gamma}_i
\end{pmatrix}
\sim
N\left(
\begin{pmatrix}
\theta_i\\
\gamma_i
\end{pmatrix},
\begin{pmatrix}
\sigma_i^2 & \rho_i\sigma_i\delta_i\\
\rho_i\sigma_i\delta_i & \delta_i^2
\end{pmatrix}
\right)$$ where

- $\sigma_i^2$ and $\delta_i^2$ are variances that reflect sampling variation
- $\rho_i$ is the correlation between the estimated treatment differences conditional upon the true differences.

Considering $\gamma_i$'s as fixed effects, we assume a simple linear model to describe the relationship between $\theta_i$ and $\gamma_i$ across the $N$ clinical trials

$$\theta_i \mid \gamma_i \sim N(\alpha + \beta\gamma_i, \tau^2).$$ By combining the within-trial distribution with the between-trial distribution, we obtain a bivariate normal distribution:

$$\begin{pmatrix}
\hat{\theta}_i\\
\hat{\gamma}_i
\end{pmatrix}
\sim
N\left(
\begin{pmatrix}
\alpha + \beta\gamma_i\\
\gamma_i
\end{pmatrix},
\begin{pmatrix}
\sigma_i^2 + \tau^2 & \rho_i\sigma_i\delta_i\\
\rho_i\sigma_i\delta_i & \delta_i^2
\end{pmatrix}
\right)$$ Note that,

$$E(\hat{\theta}_i \mid \gamma_i)
=
E\left[
E(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right]
=
E(\theta_i \mid \gamma_i)
=
\alpha + \beta\gamma_i$$

and

$$\begin{aligned}
\operatorname{Var}(\hat{\theta}_i \mid \gamma_i)
&=
E\left[
\operatorname{Var}(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right]
+
\operatorname{Var}\left(
E(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right) \\[1em]
&=
E(\sigma_i^2 \mid \gamma_i)
+
\operatorname{Var}(\theta_i \mid \gamma_i) \\[1em]
&=
\sigma_i^2 + \tau^2.
\end{aligned}$$

## Surrogacy Criteria

Daniels and Hughes described a strong surrogate relationship as one with

- $\alpha = 0$, to ensure that no treatment effect on the surrogate endpoint implies no treatment effect on the final clinical outcome
- $\beta \neq 0$, establishing a relationship between treatment effects on the surrogate and final clinical outcomes
- $\tau^2 = 0$, implying that we can predict $\theta_i$ perfectly given $\gamma_i$.

## Model Fitting

### Preparing data for STAN

```{r}
N <- nrow(df1)

STAN_data.DH <- list(
  N = N,
  theta_hat = df1$OS.ln.HR,
  gamma_hat = df1$PFS.ln.HR,
  se_theta = df1$OS.se,
  se_gamma = df1$PFS.se
)
```

### STAN Model

Here is a suitable STAN model.

```{r}
#| include: false

# use setwd() suitably
DH.model.file <- "0021__04_Daniels_Hughes_model.stan"
```

```{r}
#| include: false
cat(readLines(DH.model.file), sep = "\n")
```

As IPD are not available, assume that the $\rho_i$'s for the trials with no data available are equal to some common value $\rho$, we **assume** a positive association between `HR.OS` and `HR.PFS`, accordingly we set a priori $\rho \sim U(0, 0.999)$.

Taking a Bayesian approach, we place non-informative prior distributions on the fixed effects, $\gamma_i$ and the regression coefficients:

$$\alpha \sim N(0, A_{\alpha})$$

$$\beta \sim N(0, A_{\beta})$$

$$\gamma_i \sim N(0, A_{\gamma_i})$$

where $A_{\alpha}$, $A_{\beta}$, $A_{\gamma_i}$ are large numbers; here they are taken to be $1000$.

For the between-studies variance, $\tau^2$, a uniform prior is set on $\tau$ : $\tau \sim U(0, 2)$.

For smoother posterior sampling, $\tau$ has been transformed to log-scale and so has been the prior.

$$\text{Since } \tau \sim U(0, 2); \;\; \text{say } Y = \ln(\tau), \text{ consequently }
f_Y(y) =
\begin{cases}
\dfrac{e^y}{2}, & y < \ln 2,\\[6pt]
0, & \text{otherwise}.
\end{cases}$$

The within-trial bivariate normal distribution is represented by Cholesky decomposition of the covariance matrix, this is supposed to be more friendly with CmdstanR framework.

### Model Compilation

```{r}
DH.model <- cmdstan_model(stan_file = DH.model.file)
```

### Running HMC

```{r}
n.chains <- 3
n.samples.per.chain <- 150000

n.posterior.samples <- n.chains * n.samples.per.chain
```

```{r}
init <- function() {
  list(
    gamma = rep(0, N),
    z = rep(0, N),
    alpha = 0,
    beta = 0,
    log_tau = log(0.2),
    rho = 0.3
  )
}
```

```{r}
#| echo: true
#| output: false
fit <- DH.model$sample(
  data = STAN_data.DH,
  chains = n.chains,
  parallel_chains = n.chains,
  iter_warmup = 50000,
  iter_sampling = n.samples.per.chain,
  adapt_delta = 0.9999,
  max_treedepth = 25,
  init = init,
  refresh = 0
)
```

### STAN Diagnostics

```{r}
fit$cmdstan_diagnose()
```

These divergence is primarily because of the variance parameter $\tau^2$. $`r mean(fit[["sampler_diagnostics"]](format = "df")[["divergent__"]]) * 100`$% divergence out of $`r length(fit[["sampler_diagnostics"]](format = "df")[["divergent__"]])`$ is negligible and does not affect the parameter estimates.

## Posterior Summaries of Parameters of Interest

### Point Estimates

```{r}
fit$summary(c("alpha", "beta", "tau")) |>
  gt()
```

### Interval Estimates

```{r}
posterior.draws <- fit$draws(
  variables = c("alpha", "beta", "tau"),
  format = "matrix"
)
```

```{r}
apply(posterior.draws, 2, quantile, probs = c(0.025, 0.975))
```

## Prediction of TE on Clinical Outcome in New Trials

Now use posterior samples to calculate the predictions on new future trials, credible intervals and prediction intervals. These are the main components of **trial-level association plot**.

Create a design matrix `X` representing new studies on which we will make future predictions.

```{r}
X <- matrix(c(rep(1, 1000), new.data$PFS.ln.HR),
             nrow = 1000, ncol = 2, byrow = FALSE)
```

For each new trial in `X`, corresponding to $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ posterior samples of the parameters, there will be $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ predictions of `log.HR.OS`. Below I do a matrix operation to obtain all the $1000 \times `r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ predictions.

```{r}
DH.posterior <- X %*% t(posterior.draws[, 1:2])
```

```{r}
dim(DH.posterior)
```

Each row of `DH.posterior` represents $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ posterior draws of `log.HR.OS` *i.e.* the response variable, corresponding to a new `log.HR.PFS` of a new trial.

### Point Estimates

For a new trial, to get a prediction of `log.HR.OS`, we must calculate the mean of the posterior samples.

```{r}
DH.fit <- apply(DH.posterior, 1, mean)
```

`DH.fit` has $1000$ predictions corresponding to $1000$ new future trials.

### Credible Intervals

Credible intervals are obtained from the quantiles of the posterior distribution. Here I calculate the $95\%$ credible intervals for the predictions on new trials.

```{r}
DH.ci <- t(apply(DH.posterior, 1, quantile, probs = c(0.025, 0.975)))
```

### Prediction Intervals

Prediction intervals are obtained from the quantiles of the posterior predictive distribution.

The posterior distribution of concern here is

$$p\left(\alpha, \beta, \tau \mid Y_{11}, Y_{21}, Y_{12}, Y_{22}, \ldots, Y_{1N}, Y_{2N}\right).$$ Already I have $`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$ samples from this posterior distribution.

We want the predictive distribution of $\theta_{\text{new}}:$ the true treatment effect on clinical outcome in a future trial.

In any future trial, let $\gamma_{\text{new}}$ be the observed treatment effect on surrogate endpoint. Then we must have

$$\theta_{\text{new}} \sim \mathcal{N}\left( \alpha + \beta \cdot  \gamma_{\text{new}}, \tau^2\right).$$

In Bayesian theory, the predictive distribution is obtained by averaging the likelihood over the entire parameter space where the parameters vary in accordance to the posterior distribution.

Mathematically, we can write the posterior predictive distribution as

$$\displaystyle{\int \limits_{(\alpha, \beta, \tau)}} \mathcal{N}\left( \alpha + \beta * \gamma_{\text{new}}, \tau^2\right) \cdot p\left(\alpha, \beta, \tau \mid \cdots \right) d\alpha \; d\beta \; d\tau.$$

As analytical solution cannot be traced, we resort to numerical solution as follows :

- Using posterior sample $(\alpha_{(m)}, \beta_{(m)}, \tau_{(m)})$ calculate $\alpha_{(m)} + \beta_{(m)} * \gamma_{\text{new}}$ $\forall m = 1(1)`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$.
- Take a random sample from $\mathcal{N}\left(\alpha_{(m)} + \beta_{(m)} * \gamma_{\text{new}}, \tau_{(m)}^2 \right)$ $\forall m = 1(1)`r format(n.posterior.samples, scientific = FALSE, trim = TRUE)`$.
- Calculate $q_{0.025}$, $q_{0.975}$ for the random samples just obtained.
- Repeat the above $3$ for all the new trials to get predictive interval around their predictions.

Now I proceed to calculate the $95\%$ posterior prediction intervals.

```{r}
u <- rnorm(n.posterior.samples, 0, sd = posterior.draws[, 3])

DH.predictions <- DH.posterior +
                  matrix(rep(u, 1000), 
                         nrow = 1000, ncol = n.posterior.samples, byrow = TRUE)
```

```{r}
DH.pi <- t(apply(DH.predictions, 1, quantile, probs = c(0.025, 0.975)))
```

## Trial-level Association Plot

```{r}
DH.pred.df <- data.frame(HR.OS = exp(DH.fit),
                         HR.PFS = exp(new.data$PFS.ln.HR),
                         lower.ci = exp(DH.ci[,1]),
                         upper.ci = exp(DH.ci[,2]),
                         lower.pi = exp(DH.pi[,1]),
                         upper.pi = exp(DH.pi[,2]))
```

```{r}
DH.association.plot <- ggplot(df1, aes(x = HR.PFS, y = HR.OS)) +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.ci, ymax = upper.ci), data = DH.pred.df, alpha = 0.6, fill = "steelblue") +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.pi, ymax = upper.pi), data = DH.pred.df, alpha = 0.4, fill = "lightblue") +
  geom_line(aes(x = HR.PFS, y = HR.OS), data = DH.pred.df, 
            col = "steelblue", linewidth = 1, stat = 'identity') +
  geom_point(aes(size = weight), col = "gray30", fill = "orange", shape = 21, data = df1) +
  scale_x_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  scale_y_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  xlab("Progression-free survival hazard ratio (log scale)")+ 
  ylab(paste("Overall survival hazard ratio" , "\n", "\U00B1", "95% confidence and prediction intervals (log scale)")) +
  ggtitle("Daniels and Hughes Bayesian Bivariate Meta Analysis") +
  theme_bw() +
  scale_colour_hue("trial") +
  theme(legend.position = "none")
```

```{r}
DH.association.plot
```

## Surrogate Threshold Effect

```{r}
candidate.HR.PFS <- DH.pred.df %>%
  filter(upper.pi < 1) %>%
  pull(HR.PFS)

DH.STE <- max(candidate.HR.PFS) |> round(3)
DH.STE
```
