---
title: "Bayesian NMA — Risk Difference, Risk Ratio, NNT/NNH"
author: "Sneha"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
params:
  trt_ref: "Placebo"                   # reference/control treatment label, must match data exactly
  trt_effects: "fixed"                 # "fixed" (default) or "random" -- see note below
  chains: 3
  iter: 4000
  warmup: 2000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(multinma)   # Stan-based Bayesian NMA
library(dplyr)
library(knitr)
library(ggplot2)    # forest plot
options(mc.cores = parallel::detectCores())
set.seed(1234)
```

# About this template

This report is parameterized so the **same code runs for any binomial
outcome/treatment set** — the data is defined directly in the "Load data"
section below, no external CSV/Excel file needed.

**Fixed vs. random effects (`trt_effects`):** if each treatment contrast in
your network comes from only one study (a "star" network via a common
control, as in the example data below), a between-study heterogeneity
variance is not identifiable and `trt_effects = "fixed"` should be used. Only
switch to `"random"` once your data include multiple studies per contrast or
multi-arm trials creating closed loops.

## Expected data format

| column  | meaning                                              |
|---------|-------------------------------------------------------|
| study   | study/trial identifier                                 |
| trt     | treatment name (must include `params$trt_ref` exactly) |
| r       | number of events                                        |
| n       | arm sample size                                          |
| outcome | *(optional)* — if present, one full analysis section is produced per unique value, looping automatically |

# Load data

```{r load-data}
raw_data <- data.frame(
  outcome = c("Severe TEAE","Severe TEAE","Severe TEAE","Severe TEAE"),
  study   = c("Xu 2025","Xu 2025","Neptunus","Neptunus"),
  trt     = c("Placebo","Telitacicept_160mg","Placebo","VAY736QM"),
  r       = c(3,3,18,14),
  n       = c(127,127,307,305)
)

required_cols <- c("study","trt","r","n")
missing_cols  <- setdiff(required_cols, names(raw_data))
if (length(missing_cols) > 0) {
  stop("Data is missing required column(s): ", paste(missing_cols, collapse = ", "))
}
if (!"outcome" %in% names(raw_data)) raw_data$outcome <- "Outcome"

if (!params$trt_ref %in% raw_data$trt) {
  stop("params$trt_ref ('", params$trt_ref, "') not found in the 'trt' column of the data.")
}

kable(raw_data, caption = "Loaded arm-level data")
```

# Analysis functions

```{r functions}
run_nma <- function(df, trt_ref, trt_effects, chains, iter, warmup) {
  net <- set_agd_arm(df, study = study, trt = trt, r = r, n = n, trt_ref = trt_ref)
  fit <- nma(
    net,
    trt_effects     = trt_effects,
    link            = "logit",
    likelihood      = "binomial2",
    prior_intercept = normal(scale = 10),
    prior_trt       = normal(scale = 10),
    chains = chains, iter = iter, warmup = warmup, refresh = 0
  )
  list(net = net, fit = fit)
}

# Convert a vector of posterior log-OR draws to absolute-scale measures,
# given the baseline (comparator-arm) risk p0 that the OR is being applied to
or_to_measures <- function(logor, p0) {
  odds0 <- p0 / (1 - p0)
  odds1 <- odds0 * exp(logor)
  p1    <- odds1 / (1 + odds1)
  data.frame(RR = p1 / p0, RD = p1 - p0, NNH = 1 / abs(p1 - p0))
}

summ_row <- function(x, label, measure) {
  data.frame(
    Measure = measure, Comparison = label,
    Mean = mean(x), Lower95 = quantile(x, .025), Upper95 = quantile(x, .975),
    row.names = NULL
  )
}

# Runs the NMA for one outcome's data frame and returns OR/RR/RD/NNH for
# every treatment vs trt_ref, plus every indirect active-vs-active contrast.
#
# KEY FIX: absolute-scale measures (RR/RD/NNH) always need *some* baseline
# risk to apply the OR to. For "X vs trt_ref" that baseline is trt_ref's own
# observed risk. For an indirect "X vs Y" contrast (neither arm is trt_ref),
# the baseline must be Y's own observed risk -- not trt_ref's -- otherwise
# the OR gets projected from the wrong starting probability, which distorts
# RR/RD/NNH whenever arms have materially different baseline event rates.
analyse_outcome <- function(df, trt_ref, trt_effects, chains, iter, warmup) {
  fit_obj <- run_nma(df, trt_ref, trt_effects, chains, iter, warmup)
  net <- fit_obj$net; fit <- fit_obj$fit

  re_draws    <- as.matrix(relative_effects(fit))   # columns = non-reference treatments
  active_cols <- colnames(re_draws)

  # observed risk for EVERY treatment arm (reference included), not just trt_ref
  baseline_risks <- df %>%
    group_by(trt) %>%
    summarise(p = mean(r / n), .groups = "drop") %>%
    tibble::deframe()

  p0 <- baseline_risks[[trt_ref]]   # kept for backward compatibility / reporting

  rows <- list()

  # --- direct comparisons: X vs trt_ref -> baseline is trt_ref's own risk ---
  for (col in active_cols) {
    m <- or_to_measures(re_draws[, col], p0)
    trt_label <- sub("^d\\[(.*)\\]$", "\\1", col)
    label <- paste(trt_label, "vs", trt_ref)
    rows[[length(rows) + 1]] <- rbind(
      summ_row(exp(re_draws[, col]), label, "OR"),
      summ_row(m$RR,  label, "RR"),
      summ_row(m$RD,  label, "RD"),
      summ_row(m$NNH, label, "NNH")
    )
  }

  # --- indirect comparisons: X vs Y -> baseline is Y's OWN observed risk ---
  if (length(active_cols) >= 2) {
    for (p in combn(active_cols, 2, simplify = FALSE)) {
      logor_ind <- re_draws[, p[2]] - re_draws[, p[1]]   # logOR(t2 vs t1)
      t1 <- sub("^d\\[(.*)\\]$", "\\1", p[1])             # comparator (denominator)
      t2 <- sub("^d\\[(.*)\\]$", "\\1", p[2])             # numerator

      p0_indirect <- baseline_risks[[t1]]   # <-- t1's own risk, NOT trt_ref's
      if (is.null(p0_indirect) || is.na(p0_indirect)) {
        stop("No observed baseline risk found for '", t1,
             "' -- cannot back-transform this indirect contrast.")
      }

      m <- or_to_measures(logor_ind, p0_indirect)
      label <- paste(t2, "vs", t1)
      rows[[length(rows) + 1]] <- rbind(
        summ_row(exp(logor_ind), label, "OR"),
        summ_row(m$RR,  label, "RR"),
        summ_row(m$RD,  label, "RD"),
        summ_row(m$NNH, label, "NNH")
      )
    }
  }

  list(net = net, fit = fit, p0 = p0,
       baseline_risks = baseline_risks,
       results = do.call(rbind, rows))
}
```

# Results by outcome

Each outcome value in the data gets its own network plot, Stan model fit, and
posterior OR / RR / RD / NNH summary table below — no code changes needed to
add more outcomes, just add more rows (with a new `outcome` label) to the
data frame above.

```{r results, results='asis'}
for (oc in unique(raw_data$outcome)) {

  cat("\n\n## Outcome:", oc, "\n\n")

  df_oc <- raw_data %>% filter(outcome == oc) %>% select(study, trt, r, n)

  out <- analyse_outcome(df_oc, params$trt_ref, params$trt_effects,
                          params$chains, params$iter, params$warmup)

  cat("Baseline (", params$trt_ref, ") risk used: ",
      round(out$p0, 4),
      " (mean observed reference-arm risk across included studies)\n\n", sep = "")
  
    print(kable(
    data.frame(Treatment = names(out$baseline_risks),
               `Observed risk` = round(out$baseline_risks, 4),
               check.names = FALSE),
    caption = paste("Observed arm-level baseline risks used for back-transformation —", oc)
  ))

  print(plot(out$net) + ggplot2::ggtitle(paste("Network —", oc)))

  print(kable(out$results, digits = 2,
              caption = paste("Posterior OR / RR / RD / NNH —", oc)))
}
```

# Forest plot (Risk Ratio)

The plot below shows the posterior mean Risk Ratio (RR) and 95% credible
interval for every comparison in `out$results` (direct comparisons vs.
`Placebo`, plus the indirect active-vs-active comparison). The indirect
comparison is placed at the top of the plot, comparator labels use "vs.",
and `Placebo` is abbreviated "PBO".

```{r rr-table, include=FALSE}
RR_table <- out$results %>%
  filter(Measure == "RR") %>%
  transmute(Comparison = Comparison, RR = Mean, Lower = Lower95, Upper = Upper95)
```

```{r forest-data-prep, include=FALSE}
comparison_labels <- gsub("_", " ", RR_table$Comparison)
comparison_labels <- gsub("\\bvs\\b", "vs.", comparison_labels)
comparison_labels <- gsub("\\bPlacebo\\b", "PBO", comparison_labels)

df1 <- data.frame(comparison = comparison_labels,
                   RR = RR_table$RR,
                   lower = RR_table$Lower,
                   upper = RR_table$Upper)
```

```{r forest-data-flag, include=FALSE}
df2 <- df1 %>%
  mutate(
    Significant = dplyr::if_else(lower < 1 & upper < 1, "Yes", "No"),
    Label = sprintf("%.2f (%.2f, %.2f)", RR, lower, upper)
  ) %>%
  # Preserve the row order from RR_table (direct comparisons first, indirect
  # last) -- with ggplot's default discrete-axis mapping this places the
  # indirect comparison at the TOP of the forest plot.
  mutate(comparison = factor(comparison, levels = unique(comparison)))
```

```{r forest-plot-build, include=FALSE}
forest_plot <- df2 %>% ggplot(
  aes(x = RR,
      y = comparison,
      xmin = lower,
      xmax = upper,
      colour = Significant)) +
  geom_errorbar(orientation = "y", width = 0.25, linewidth = 0.9, colour = "#2166AC") +
  geom_point(size = 7, shape = 18, colour = "#2166AC") +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey40", linewidth = 0.7) +
  geom_text(aes(x = upper, label = Label),
            hjust = -0.08, size = 6, colour = "grey20") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.35))) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey40", size = 10),
        plot.caption = element_text(colour = "grey50", size = 9),
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(size = 15))
```

```{r forest-plot-show, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
forest_plot
```

```{r forest-plot-save, include=FALSE, message=FALSE, warning=FALSE}
ggsave("FP_SevereTEAE.png",
       forest_plot,
       width = 11.5,
       height = 5.5,
       dpi = 2000,
       bg = "white",
       create.dir = TRUE)
```

# Notes

- **RD sign / NNH:** `RD > 0` means higher event risk than the comparator.
  Measures here are labelled NNH (number needed to harm) since the outcome
  is an adverse event; relabel as NNT if you reuse this for a benefit
  outcome.
- **Baseline risk:** absolute measures (RD, RR, NNH) depend on the chosen
  baseline risk — this template uses the mean *observed* reference-arm risk
  per outcome. Swap in an external/trial-specific value in `analyse_outcome()`
  if your SAP specifies one.
- **Indirect comparisons:** for outcomes with more than one active treatment,
  all pairwise contrasts between active treatments are derived automatically
  from the NMA-estimated effects vs. the common reference (fully Bayesian,
  Bucher-equivalent).
