---
title: "Frequentist Weighted Linear Regression"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Required Libraries

```{r message=FALSE, warning=FALSE}
library(openxlsx)

library(wCorr)
library(DescTools)

library(gt)
library(tidyverse)
library(cmdstanr)
```

## Loading Data

```{r}
#| include: false
data.path <- 'https://github.com/ananda-ebm/datasets/raw/refs/heads/master/NSCLC.xlsx'
```

```{r}
df1 <- read.xlsx(data.path, sheet = 1)
```

## Initial Sanity Checks

```{r}
str(df1)
```

```{r}
dim(df1)
```

```{r, results='asis'}
#| echo: false
#| tbl-cap: "Snapshot of NSCLC Data"

head(df1) |>
  gt() |>
  tab_options(
    data_row.padding = px(8),
    column_labels.padding = px(10)
  )
```

## Generating Treatment Effect on Surrogate Endpoint in New Trials

```{r}
range(df1$HR.PFS)
```

```{r}
# Suppose 1000 new trials reporting TE on surrogate endpoint only
new.data <- data.frame(PFS.ln.HR = log(seq(0.2,3, length=1000)))
```

## Some Derived Quantities

The weights used in this analysis were the inverse variance $1/{\text{SE}^2}$ of the log HR of OS.

```{r}
df1$weight <- 1 / (df1$OS.se)^2
```

The weighted correlation coefficient $(\rho)$ was estimated using the Pearson weighted correlation of the target and the surrogate endpoints.

```{r}
rho <- weightedCorr(df1$OS.ln.HR, df1$PFS.ln.HR,
             method = "Pearson",
             weights = 1 / df1$OS.se^2)
rho
```

Confidence intervals around Pearson $\rho$ are not symmetrical. Therefore, a Fisher's $z-$transformation was applied to estimate the $95\%$ confidence intervals, which is conducted using the `CorCI()` function in the `DescTools` package.

```{r}
effective.n <- sum(df1$weight)^2 / sum(df1$weight^2)
effective.n <- round(effective.n, 1)

effective.n
```

```{r}
CorCI(rho = rho, 
      n = effective.n, 
      conf.level = 0.95, alternative = "two.sided")
```

## Model

The weighted linear regression (WLR) model is given by

$$\theta_i = \alpha + \beta\gamma_i + \varepsilon_i,
\qquad i = 1,\ldots,n,$$

where

$$\varepsilon_i \sim
N\left(0,\sigma_{\theta_i}^2\right); 
\qquad 
w_i = \dfrac{1}{\sigma_{\theta_i}^2}.$$

The parameters $\alpha$ and $\beta$ are estimated by minimizing the weighted sum of squared residuals:

$$\underset{\alpha,\beta}{\operatorname{minimize}}
\sum_{i=1}^{n}
w_i
\left(
\theta_i-\alpha-\beta\gamma_i
\right)^2.$$ Here:

- $\theta_i$ is the treatment effect on the final endpoint in study $i$;
- $\gamma_i$ is the treatment effect on the surrogate endpoint in study $i$;
- $\alpha$ is the intercept;
- $\beta$ is the regression coefficient measuring the association between treatment effects on the surrogate and final endpoints; and
- $w_i$ is the weight assigned to study $i$.

## Model Fitting

Frequentist weighted linear regression is arguably the *simplest* approach to a surrogate analysis. A weighted linear regression model was used to relate the target endpoint (log HRs of OS) to the surrogate endpoint (log HRs of PFS). The data were *weighted* by the inverse of the variance of the log HRs of the target variable.

```{r}
fit1 <- lm(OS.ln.HR ~ PFS.ln.HR, weights = weight, data = df1)
summary(fit1)
```

```{r}
sqrt(summary(fit1)$r.squared)
```

## Prediction of TE on Clinical Outcome in New Trials

### Confidence Intervals

```{r}
wlr.ci <- data.frame(predict(fit1, newdata = new.data, interval = "confidence"))

head(wlr.ci)
```

### Prediction Intervals

```{r}
wlr.pi <- data.frame(predict(fit1, newdata = new.data, interval = "prediction",
                  weight = 1 / (mean(df1$OS.se)^2) ))

head(wlr.pi)
```

## Trial-level Association Plot

```{r}
wlr.pred.df <- data.frame(HR.OS = exp(wlr.ci[,1]),
                        HR.PFS = exp(new.data$PFS.ln.HR),
                        lower.ci = exp(wlr.ci[,2]),
                        upper.ci = exp(wlr.ci[,3]),
                        lower.pi = exp(wlr.pi[,2]),
                        upper.pi = exp(wlr.pi[,3]))
```

```{r}
wlr.association.plot <- ggplot(df1, aes(x = HR.PFS, y = HR.OS)) +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.ci, ymax = upper.ci), data = wlr.pred.df, stat = 'identity', fill = "steelblue", alpha = 0.5) +
  geom_ribbon(aes(x = HR.PFS, ymin = lower.pi, ymax = upper.pi), data = wlr.pred.df, stat = 'identity', fill = "lightblue", alpha = 0.4) +
  scale_x_continuous(trans = "log", 
                     limits = c(min(df1$HR.PFS)*0.9, max(df1$HR.PFS)*1.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  scale_y_continuous(trans = "log",
                     limits=c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  geom_smooth(aes(x = HR.PFS, y = HR.OS), data = wlr.pred.df, stat = 'identity') +
  geom_point(aes(x = HR.PFS, y = HR.OS, size = weight), 
             col = "gray30", shape = 21, fill = "orange") +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  xlab("Progression-free survival hazard ratio (log scale)")+ 
  ylab(paste("Overall survival hazard ratio" , "\n", "\U00B1", "95% confidence and prediction intervals (log scale)")) +
  ggtitle("Weighted Linear Regression") +
  theme_bw() +
  scale_colour_hue("trial") +
  theme(legend.position = "none")
```

```{r, fig.width=8, fig.height=6, out.width="100%"}
wlr.association.plot
```

## Surrogate Threshold Effect

Surrogate Threshold Effect is defined as ***the minimum treatment effect on the surrogate necessary to predict a nonzero effect on the true endpoint*** in a future trial. $\log(HR) < 0$ implies a *beneficial* treatment effect. So STE was computed as the intersection of the upper $95\%$ prediction interval with the horizontal line for no treatment effect (i.e., in which the HR for OS equals $1$ or $\log(HR) = 0$).

```{r}
candidate.HR.PFS <- wlr.pred.df %>%
  filter(upper.pi < 1) %>%
  pull(HR.PFS)

WLR.STE <- round(max(candidate.HR.PFS), digits = 3)

WLR.STE
```
