---
title: "Frequentist Fixed-effects Meta Regression"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-tools: true
    code-summary: "Show/Hide Code"
knitr:
  opts_chunk:
    fig.path: "figures/"
---

## Required Libraries

```{r message=FALSE, warning=FALSE}
library(openxlsx)

library(gt)
library(tidyverse)

library(metafor)
```

## Loading Data

```{r}
#| include: false
data.path <- 'https://github.com/ananda-ebm/datasets/raw/refs/heads/master/NSCLC.xlsx'
```

```{r}
df1 <- read.xlsx(data.path, sheet = 1)
```

## Initial Sanity Checks

```{r}
str(df1)
```

```{r}
dim(df1)
```

```{r, results='asis'}
#| echo: false
#| tbl-cap: "Snapshot of NSCLC Data"

head(df1) |>
  gt() |>
  tab_options(
    data_row.padding = px(8),
    column_labels.padding = px(10)
  )
```

## Derived Quantities

```{r}
df1$weight <- 1 / (df1$OS.se)^2
```

## Generating Treatment Effect on Surrogate Endpoint in New Trials

```{r}
range(df1$HR.PFS)
```

```{r}
# Suppose 1000 new trials reporting TE on surrogate endpoint only
new.data <- data.frame(PFS.ln.HR = log(seq(0.2,3, length=1000)))
```

## Model

Meta-regression is an extension of meta-analysis that includes the surrogate endpoint as a study-level covariate. The model is written as

$$\widehat{\theta}_i
=
\theta+\beta\widehat{\alpha}_i+\varepsilon_i,
\qquad
i=1,\ldots,N,$$

where

$$\varepsilon_i
\overset{\mathrm{iid}}{\sim}
N(0,\sigma^2).$$ Here:

- $\widehat{\theta}_i$ = estimated treatment effect on the final endpoint in study $i$;
- $\widehat{\alpha}_i$ = estimated treatment effect on the surrogate endpoint in study $i$;
- $\theta$ = intercept;
- $\beta$ = regression coefficient describing the association between surrogate and final treatment effects.

Here we assume that $\theta$ is common for all studies and the only reason observed treatment effect $\widehat{\theta_i}$'s are different from $\theta$ is the sampling error (in addition to the observed treatment effect on surrogate endpoint of that study).

## Model Fitting

```{r}
fit.FEMR <- metafor::rma(OS.ln.HR ~ 1 + PFS.ln.HR, sei = OS.se, data = df1, method = "FE")
summary(fit.FEMR)
```

## Prediction of TE on Clinical Outcome in New Trials and Confidence Intervals

We assume that there is no between-trial heterogeneity in either the target or the surrogate endpoint (fixed-effects assumption). So confidence intervals can be interpreted as prediction intervals.

```{r}
FEMR.ci <- data.frame(predict(fit.FEMR, new.data$PFS.ln.HR))
head(FEMR.ci) |>
  gt()
```

```{r}
FEMR.ci$HR.OS <- exp(FEMR.ci$pred)
FEMR.ci$HR.PFS <- exp(new.data$PFS.ln.HR)

head(FEMR.ci) |>
  gt()
```

## Trial-level Association Plot

```{r}
FEMR.association.plot <- ggplot(df1, aes(x = HR.PFS, y = HR.OS)) +
  geom_ribbon(aes(x = HR.PFS, ymin = exp(ci.lb), ymax = exp(ci.ub)), data = FEMR.ci, fill = "steelblue", alpha = 0.4) +
  geom_smooth(aes(x = HR.PFS, y = HR.OS), data = FEMR.ci, stat = 'identity') +
  geom_point(aes(size = weight), shape = 21, col = "gray30", fill = "orange") +
  scale_x_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  scale_y_continuous(transform = "log",
                     limits = c(0.2, 3.1),
                     breaks=c(0.1,0.2,0.3,0.4,0.5,0.7,1,1.2,1.5,2,3),
                     expand = c(0.01, 0.01)) +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  xlab("Progression-free survival hazard ratio (log scale)")+ 
  ylab(paste("Overall survival hazard ratio" , "\n", "\U00B1", "95% confidence and prediction intervals (log scale)")) +
  ggtitle("Frequentist Fixed-effects Meta-regression") +
  theme_bw() +
  scale_colour_hue("trial") +
  theme(legend.position = "none")
```

```{r, fig.width=8, fig.height=6, out.width="100%"}
FEMR.association.plot
```

## Surrogate Threshold Effect

```{r}
candidate.HR.PFS <- FEMR.ci %>%
  filter(ci.ub < 0) %>%
  pull(HR.PFS)

FEMR.STE <- round(max(candidate.HR.PFS), digits = 3)
FEMR.STE
```
