---
title: "Daniels and Hughes Bivariate Meta-analytic Method for Evaluation of Potential Surrogate Markers"
author: "Ananda Biswas"
date: "`r Sys.Date()`"
bibliography: references.bib
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-tools: true
    code-summary: "Show/Hide Code"
knitr:
  opts_chunk:
    fig.path: "figures/"
---

## Model

Consider a meta-analysis of $N$ clinical trials each of which, for simplicity, involves the comparison of two treatments.

In the $i-$th trial,

- $\theta_i$ denotes the true treatment difference on the clinical outcome (true endpoint),
- $\gamma_i$ denotes the true treatment difference on the response variable (surrogate endpoint).

We assume that we have estimates $\widehat{\theta_i}$ and $\widehat{\gamma_i}$, respectively, available from each trial and that the size of each study is sufficiently large, that $(\widehat{\theta_i}, \widehat{\gamma_i})$ is normally distributed about $(\theta_i, \gamma_i)$.

Within the $i-$th clinical trial, Daniels and Hughes [@danielshughes1997] assume that the following model applies:

$$\begin{pmatrix}
\hat{\theta}_i\\
\hat{\gamma}_i
\end{pmatrix}
\mid
\theta_i,
\gamma_i
\sim
N\left(
\begin{pmatrix}
\theta_i\\
\gamma_i
\end{pmatrix},
\begin{pmatrix}
\sigma_i^2 & \rho_i\sigma_i\delta_i\\
\rho_i\sigma_i\delta_i & \delta_i^2
\end{pmatrix}
\right)$${#eq-within-trial-distribution} where

- $\sigma_i^2$ and $\delta_i^2$ are variances that reflect sampling variation
- $\rho_i$ is the correlation between the estimated treatment differences *conditional upon the true differences*.

Considering $\gamma_i$'s as **fixed effects**, we assume a simple linear model to describe the relationship between $\theta_i$ and $\gamma_i$ across the $N$ clinical trials

$$\theta_i \mid \gamma_i \sim N(\alpha + \beta\gamma_i, \tau^2).$${#eq-between-trials-distribution}

By combining the within-trial distribution [-@eq-within-trial-distribution] with the between-trials distribution [-@eq-between-trials-distribution], we obtain a bivariate normal distribution:

$$\begin{pmatrix}
\hat{\theta}_i\\
\hat{\gamma}_i
\end{pmatrix}
\sim
N\left(
\begin{pmatrix}
\alpha + \beta\gamma_i\\
\gamma_i
\end{pmatrix},
\begin{pmatrix}
\sigma_i^2 + \tau^2 & \rho_i\sigma_i\delta_i\\
\rho_i\sigma_i\delta_i & \delta_i^2
\end{pmatrix}
\right)$${#eq-combined-distribution}

Note that,

$$E(\hat{\theta}_i \mid \gamma_i)
=
E\left[
E(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right]
=
E(\theta_i \mid \gamma_i)
=
\alpha + \beta\gamma_i$$

and

$$\begin{aligned}
\operatorname{Var}(\hat{\theta}_i \mid \gamma_i)
&=
E\left[
\operatorname{Var}(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right]
+
\operatorname{Var}\left(
E(\hat{\theta}_i \mid \theta_i, \gamma_i)
\mid \gamma_i
\right) \\[1em]
&=
E(\sigma_i^2 \mid \gamma_i)
+
\operatorname{Var}(\theta_i \mid \gamma_i) \\[1em]
&=
\sigma_i^2 + \tau^2.
\end{aligned}$$

## Surrogacy Criteria

A strong surrogate relationship [@TSD20] is one with

- $\alpha = 0$, to ensure that no treatment effect on the surrogate endpoint implies no treatment effect on the final clinical outcome
- $\beta \neq 0$, establishing a relationship between treatment effects on the surrogate and final clinical outcomes
- $\tau^2 = 0$, implying that we can predict $\theta_i$ perfectly given $\gamma_i$.

In practice, we consider a surrogate relationship strong if all the following is true:

- the $95\%$ interval for the intercept $\alpha$ includes zero
- the $95\%$ interval for the slope $\beta$ does not include zero
- the conditional variance $\tau^2$ along with its upper interval is small.

## Required Packages

```{r}
#| message: false
#| warning: false
library(MASS)
library(gt)
library(cmdstanr)
```

```{r}
#| message: true
#| warning: true
#| include: false
options(mc.cores = parallel::detectCores())
```

## Simulation

```{r}
set.seed(14)
```

In our simulation study, we generate data from a scenario where surrogacy criteria are satisfied. We set number of trials $N = 30$; intercept $\alpha = 0$; slope $\beta = 0.8$ and conditional variance $\tau^2$ is close to $0$.

```{r}
N <- 30

alpha.true <- 0.0; beta.true <- 0.8

tau.true <- 0.1
```

True treatment effects on surrogate endpoint $\gamma_i$'s are fixed effects, we generate $N$ of such fixed effects.

```{r}
gamma.true <- rnorm(N, mean = 3, sd = 1.5)
```

Following between-trials distribution [-@eq-between-trials-distribution], we generate true treatment effect on true endpoint $\theta_i$'s.

```{r}
theta.true <- rnorm(N,
                    mean = alpha.true + beta.true * gamma.true,
                    sd = tau.true)
```

We choose some arbitrary values for within-trial variances and covariances.

```{r}
sigma.true <- runif(N, 0.1, 0.5)
delta.true <- runif(N, 0.1, 0.7)
rho.true <- runif(N, -0.2, 0.2)
```

Construct within-trial covariance matrices as follows.

```{r}
Sigma <- array(0, dim = c(N, 2, 2))

for (i in 1:N) {
  Sigma[i, 1, 1] <- sigma.true[i]^2
  Sigma[i, 1, 2] <- rho.true[i] * sigma.true[i] * delta.true[i]
  Sigma[i, 2, 1] <- rho.true[i] * sigma.true[i] * delta.true[i]
  Sigma[i, 2, 2] <- delta.true[i]^2
}
```

Following within-trial distribution [-@eq-within-trial-distribution], we simulate observed treatment effects on true endpoint and surrogate endpoint as follows.

```{r}
theta.hat <- gamma.hat <- c()

for (i in 1:N) {
  
  obs <- MASS::mvrnorm(1,
                       mu = c(theta.true[i], gamma.true[i]),
                       Sigma = Sigma[i, , ])
  
  theta.hat[i] <- obs[1]
  gamma.hat[i] <- obs[2]
}
```

In this simulation study, we **assume** that within-trial correlations are also included in our data. We combine all the simulated data in a data-frame.

```{r}
df <- data.frame(
  study = 1:N,
  theta.hat = theta.hat,
  se.theta = sigma.true,
  gamma.hat = gamma.hat,
  se.gamma = delta.true,
  rho.w = rho.true
)
```

```{r}
df |> gt()
```

## Model Fitting

### Preparing data for STAN

```{r}
DH.data <- list(
  N = nrow(df),
  theta_hat = df$theta.hat,
  gamma_hat = df$gamma.hat,
  se_theta = df$se.theta,
  se_gamma = df$se.gamma,
  rho_w = df$rho.w
)
```

### STAN Model

For evaluation of candidate surrogate endpoint, the parameters of interest are $\alpha, \beta$ and $\tau^2$; in the sense that we require posterior estimates of these parameters only. 

Taking a Bayesian approach, we place non-informative prior distributions on the fixed effects, $\gamma_i$ and the regression coefficients:

$$\alpha \sim N(0, A_{\alpha})$$

$$\beta \sim N(0, A_{\beta})$$

$$\gamma_i \sim N(0, A_{\gamma_i})$$

where $A_{\alpha}$, $A_{\beta}$, $A_{\gamma_i}$ are large numbers; here they are taken to be $1000$.

For the between-studies variance, $\tau^2$, a uniform prior is set on $\tau$ : $\tau \sim U(0, 2)$.

For smoother posterior sampling, $\tau$ has been transformed to log-scale and so has been the prior.

$$\text{Since } \tau \sim U(0, 2); \;\; \text{say } Y = \ln(\tau), \text{ consequently }
f_Y(y) =
\begin{cases}
\dfrac{e^y}{2}, & y < \ln 2,\\[6pt]
0, & \text{otherwise}.
\end{cases}$$

Following the combined distribution [-@eq-combined-distribution], here is a suitable STAN model.

```{r}
DH.model.file <- "Daniels_and_Hughes_Bivariate_Meta_Analysis_01.stan"
```

```{r}
#| echo: false
#| results: asis
cat("```stan\n", paste(readLines(DH.model.file), collapse = "\n"), "\n```", sep = "")
```

### Model Compilation

```{r}
#| message: false
#| warning: false
DH.model <- cmdstan_model(DH.model.file)
```

### Running HMC

```{r}
n.chains <- 3
n.samples.per.chain <- 150000

n.posterior.samples <- n.chains * n.samples.per.chain
```

```{r}
fit <- DH.model$sample(
  data = DH.data,
  chains = n.chains,
  parallel_chains = n.chains,
  iter_warmup = 50000,
  iter_sampling = n.samples.per.chain,
  adapt_delta = 0.95,
  max_treedepth = 15,
  seed = 14,
  refresh = 0
)
```

### HMC Diagnostics

```{r}
fit$cmdstan_diagnose()
```

## Posterior Summaries of Parameters of Interest

```{r}
fit$print(variables = c("alpha", "beta", "tau"))
```

Posterior mean and median of the parameters firmly agree with the true values of the respective parameters, indicating **praise-worthy parameter recovery**.

## Assessing Surrogacy

Let us have posterior draws of the parameters of interest, so that we can obtain their $95\%$ credible intervals.

```{r}
posterior.draws <- fit$draws(variables = c("alpha", "beta", "tau_sq"), 
                             format = "matrix")
```

```{r}
apply(posterior.draws, 2, quantile, probs = c(0.025, 0.975)) |> round(5)
```

- The $95\%$ credible interval for the intercept $\alpha$ includes zero.
- The $95\%$ credible interval for the slope $\beta$ does not include zero.
- The conditional variance $\tau^2$ along with its upper interval is small.

Hence, trial-level association of the candidate surrogate endpoint is *validated*.